---
title: "curate tax assignments and decon ASVs"
output: html_document
date: "2026-04-13"
author: "Kimberly Ledger"
---

```{r}
library(tidyverse)
library(ggplot2)
library(here)
```

perform additional taxonomic curation and summary of ASV taxonomic assignments
```{r}
## load taxonomic assignments
assignments <- read.csv(here("data", "collapsed_taxonomy.csv")) %>% unique() %>% select(!X)

## load LULU output
map <- read.csv(here("data", "lulu_ref", "asvmap_lulu98.csv")) %>%
  rename(ASV = X)

## merge 
join <- map %>%
  left_join(assignments, by = "ASV") %>%
  rename(ASV_taxon = taxon) %>%
  select(ASV:curated, ASV_taxon) %>%
  left_join(assignments, by = c("parent_id" = "ASV")) %>%
  rename(parent_taxon = taxon) %>%
  select(ASV:ASV_taxon, parent_taxon)

mismatches <- join %>%
  filter(ASV_taxon != parent_taxon) %>%
  filter(!str_detect(parent_taxon, " "))
## if parent taxa is species level, then i will go with that ID 

## if ASV taxa is species level and parent ID is not, then go with original ASV ID
## except with that particular species is not in the dataset 
mismatches_species <- mismatches %>%
  filter(str_detect(ASV_taxon, " ")) %>% 
  filter(!ASV_taxon %in% c("Anoplarchus purpurescens", "Sebastes babcocki"))

keeper_1 <- mismatches_species$ASV
## otherwise, use the parent ASV id  

# just checking these 
mismatches_notspecies <- mismatches %>%
  filter(!str_detect(ASV_taxon, " ")) 
```

update asv to use for taxonomic assignment 
```{r}
map_updated <- map %>%
  select(!spread) %>%
  select(!rank) %>%
  mutate(parent_id = ifelse(ASV %in% keeper_1, ASV, parent_id),
         curated = ifelse(ASV %in% keeper_1, "updated", curated))
```

merge taxonomic assignments with ASV table 
```{r}
asv_table <- readRDS(here("data/dadasnake_20241106/filtered.seqTab.RDS")) %>%
  select(!Row.names)

#transpose 
asv_table <- data.frame(t(asv_table))
#set column names to be ASV# 
colnames(asv_table) <- asv_table["ASV",]
#remove row that has ASV#
asv_table <- asv_table[!rownames(asv_table) %in% c('ASV'), ]
#make sure reads are numbers
# Convert all character columns to numeric
for (col in names(asv_table)) {
  asv_table[[col]] <- as.numeric(asv_table[[col]])
}
#make make sample ID a column 
asv_table$Sample_ID <- rownames(asv_table)
#reorder
asv_table <- asv_table %>%
  select(Sample_ID, everything())
```

# pivot longer and join with tax assignments
```{r}
asv_table_long <- asv_table %>%
  pivot_longer(cols = -1, names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(map_updated, by = "ASV") %>%
  select(!total) %>%
  select(!curated) %>%
  rename(ASV_for_tax_ID = parent_id) %>%
  left_join(assignments, by = c("ASV_for_tax_ID" = "ASV")) %>%
##need to fix ASV_0022 & ASV_0056 = Liparis greeni
  mutate(taxon = ifelse(ASV_for_tax_ID %in% c("ASV_0022", "ASV_0056"), "Liparis greeni", taxon),
         taxonomic_level = ifelse(ASV_for_tax_ID %in% c("ASV_0022", "ASV_0056"), "species", taxonomic_level),
         species = ifelse(ASV_for_tax_ID %in% c("ASV_0022", "ASV_0056"), "Liparis greeni", species))
```


### start decontamination steps ### 

#### 1 - remove ASVs that do not have a taxonomic assingment 
```{r}
asv_table_filter1 <- asv_table_long %>%
  filter(!is.na(taxon))

removed_asvs <- anti_join(asv_table_long, asv_table_filter1) %>%
  group_by(ASV) %>%
  summarize(total_reads = sum(reads)) %>%
  arrange(desc(total_reads))
removed_asvs

#how many ASVs and total reads removed because of no tax assignment? 
nrow(removed_asvs)
sum(removed_asvs$total_reads)
#proportion of reads removed because of no taxonomic assignment
sum(removed_asvs$total_reads)/sum(asv_table$reads)
```


look at non-fish taxonomic assignments 
```{r}
asv_table_filter1 %>% filter(class != "Actinopteri") %>% filter(reads > 0) %>%
  group_by(taxon) %>%
  summarize(total_reads = sum(reads),
            n_pcr_reps = n_distinct(Sample_ID))
```

remove non-fish taxonomic assignments
```{r}
asv_table_filter2 <- asv_table_filter1 %>%
  filter(!class != "Actinopteri") 
```

load sample metadata
```{r}
metadata <- read.csv(here("data/GOA2021_metadata_20241107.csv"))

#illumina output changed "_" to "-"
metadata$Sample_ID <- gsub("_", "-", metadata$Sample_ID) 
```


add column to the ASV table that labels the sample type
```{r}
asv_table_with_sample_type <- metadata %>%
  dplyr::select(Sample_ID, run, sample_type) %>%
  left_join(asv_table_filter2, by = "Sample_ID")
```

#### 2 - account for tag-jumping using the positive controls 

first, plot
```{r}
asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(sample_type == "positive_control") %>%
  ggplot(aes(x=Sample_ID, y=reads, fill=ASV)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  facet_wrap(~run, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(sample_type == "positive_control") %>%
  ggplot(aes(x=Sample_ID, y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  facet_wrap(~run, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

identify the maximum proportion of reads for each ASV found in the positive controls
```{r}
prop_asvs_in_positives <- asv_table_with_sample_type %>%
  filter(reads > 0) %>%
  filter(sample_type == "positive_control") %>%
  group_by(Sample_ID) %>%
  mutate(TotalReadsPerSample = sum(reads)) %>%
  mutate(Prop = reads/TotalReadsPerSample) %>%
  group_by(run, ASV) %>%
  summarise(max_prop = max(Prop)) %>%
  arrange(desc(max_prop)) 
prop_asvs_in_positives
```

subtract the max proportion of tag-jumped reads for each ASV from samples
```{r}
indexhop_table <- asv_table_with_sample_type %>%
  group_by(Sample_ID) %>%
  mutate(TotalReadsPerSample = sum(reads, na.rm = T)) %>%
  left_join(prop_asvs_in_positives, by = c("ASV", "run")) %>%
  mutate(IndexHoppingReads = TotalReadsPerSample*max_prop) %>%
  mutate(IndexHoppingReads = if_else(is.na(IndexHoppingReads), 0, IndexHoppingReads)) %>%
  mutate(IndexHoppingReads = as.integer(IndexHoppingReads)) %>%
  mutate(reads_IndexHop_removed = reads - IndexHoppingReads) %>%
  mutate(reads_IndexHop_removed = if_else(reads_IndexHop_removed < 0, 0, reads_IndexHop_removed))
head(indexhop_table)
```

clean up the table by removing columns no longer needed 
```{r}
asv_table_filter3 <- indexhop_table %>%
  dplyr::select(!reads) %>%
  dplyr::select(!TotalReadsPerSample) %>%
  dplyr::select(!max_prop) %>%
  dplyr::select(!IndexHoppingReads) %>%
  dplyr::rename(reads = reads_IndexHop_removed)
```


### 3 - Account for contaminants in negative controls 

plot negatives 

first, pcr negatives 
```{r}
asv_table_filter3 %>%
  filter(reads > 0) %>%
  filter(sample_type == "PCR_blank") %>%
  ggplot(aes(x=Sample_ID, y=reads, fill=ASV)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASVs in PCR negative controls") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    #axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_filter3 %>%
  filter(reads > 0) %>%
  filter(sample_type == "PCR_blank") %>%
  ggplot(aes(x=Sample_ID, y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASVs in PCR negative controls") + 
  facet_wrap(~run, scales = "free") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    #axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank() 
  ) + 
  guides(fill = guide_legend(ncol = 2))
```

subtract the mean number of reads per ASV in the PCR negatives by sequencing run 
```{r}
mean_asvs_in_neg<- asv_table_filter3 %>%
  filter(sample_type %in% c("PCR_blank")) %>%
  group_by(ASV, run) %>%
  summarize(MeanReadsPerASV = mean(reads))
mean_asvs_in_neg

#subtract the mean number of reads for each ASV in the negative PCR controls
control_table <- asv_table_filter3 %>%
  left_join(mean_asvs_in_neg, by = c("ASV", "run")) %>%
  mutate(MeanReadsPerASV = as.integer(MeanReadsPerASV )) %>%
  mutate(MeanReadsPerASV = ifelse(is.na(MeanReadsPerASV), 0, MeanReadsPerASV)) %>%
  mutate(reads_control_removed = reads - MeanReadsPerASV) %>%
  mutate(reads_control_removed = if_else(reads_control_removed < 0, 0, reads_control_removed))
head(control_table)
```

clean up the table by removing columns no longer needed 
```{r}
asv_table_filter4 <- control_table %>%
  dplyr::select(!reads) %>%
  dplyr::select(!MeanReadsPerASV) %>%
  dplyr::rename(reads = reads_control_removed) %>%
  filter(reads > 0)
```


second, extraction negatives 
```{r}
asv_table_filter4 %>%
  filter(reads > 0) %>%
  filter(Sample_ID != "e00562-A") %>%  
  filter(sample_type == "extraction_blank") %>%
  ggplot(aes(x=Sample_ID, y=reads, fill=ASV)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASVs in extraction negative controls") + 
  facet_wrap(~run, scales = "free") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    #axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_filter4 %>%
  filter(reads > 0) %>%
  filter(Sample_ID != "e00562-A") %>%                 ## lots of reads in this extraction blank
  filter(sample_type == "extraction_blank") %>%
  ggplot(aes(x=Sample_ID, y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASVs in extraction negative controls") + 
  facet_wrap(~run, scales = "free") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    #axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank() 
  ) + 
  guides(fill = guide_legend(ncol = 2))
```

subtract the mean number of reads per ASV in the PCR negatives by sequencing run 
```{r}
mean_asvs_in_neg_ext<- asv_table_filter4 %>%
  filter(Sample_ID != "e00562-A") %>% 
  filter(sample_type %in% c("extraction_blank")) %>%
  group_by(ASV, run) %>%
  summarize(MeanReadsPerASV = mean(reads))
mean_asvs_in_neg_ext

#subtract the mean number of reads for each ASV in the negative PCR controls
control_table_ext <- asv_table_filter4 %>%
  left_join(mean_asvs_in_neg, by = c("ASV", "run")) %>%
  mutate(MeanReadsPerASV = as.integer(MeanReadsPerASV )) %>%
  mutate(MeanReadsPerASV = ifelse(is.na(MeanReadsPerASV), 0, MeanReadsPerASV)) %>%
  mutate(reads_control_removed = reads - MeanReadsPerASV) %>%
  mutate(reads_control_removed = if_else(reads_control_removed < 0, 0, reads_control_removed))
head(control_table_ext)
```

clean up the table by removing columns no longer needed 
```{r}
asv_table_filter5 <- control_table_ext %>%
  dplyr::select(!reads) %>%
  dplyr::select(!MeanReadsPerASV) %>%
  dplyr::rename(reads = reads_control_removed) %>%
  filter(reads > 0)
```

### check if any asv's are only in controls 
```{r}
asv_table_filter5 %>%
  group_by(ASV, sample_type) %>%
  summarize(TotalReadsPerASV = sum(reads, na.rm = TRUE)) %>%
  arrange(ASV) %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
    filter(sample < 1)
```

#### 4 - Remove low read depth samples based on ASV accumulation curve

```{r}
library(vegan)

asv_table_wide <- asv_table_filter5 %>%
  select(Sample_ID, ASV_for_tax_ID, reads) %>%
  filter(!is.na(reads)) %>%
  mutate(reads = as.integer(reads)) %>%
  group_by(Sample_ID, ASV_for_tax_ID) %>%
  summarise(reads = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = ASV_for_tax_ID, values_from = reads, values_fill = 0)

Sample_ID <- asv_table_wide$Sample_ID

asv_table_wide <- asv_table_wide %>%
  ungroup() %>%
  select(-Sample_ID)

## plots the figure
rarecurve(asv_table_wide, step = 20, col = "blue", label = FALSE, 
          main = "Sequencing Effort Curves",
          xlab = "Sequencing Depth", ylab = "Number of ASVs Identified",
          xlim = c(0,3000))
```


summarize in a table how many pcr replicates meet certain read count thresholds 
```{r}
read_summary <- asv_table_filter5 %>%
  filter(!is.na(reads)) %>%
  group_by(Sample_ID, sample_type) %>%
  summarize(tot_reads = sum(reads)) %>%
  arrange(desc(tot_reads)) %>%
  group_by(sample_type) %>%
  summarize(atleast0 = sum(tot_reads >= 0),
            atleast100 = sum(tot_reads >= 100),
            atleast500 = sum(tot_reads >= 500),
            atleast1k = sum(tot_reads >= 1000),
            atleast2k = sum(tot_reads >= 2000))
read_summary
```


apply a threshold of 1000 reads per pcr replicate
```{r}
asv_table_filter6 <- asv_table_filter5 %>%
  group_by(Sample_ID) %>%
  mutate(tot_reads = sum(reads)) %>%
  filter(tot_reads >= 1000)
```

```{r}
#saveRDS(asv_table_filter6, file = here("data", "decontaminated_asv_table.RDS"))
```

