---
title: "salmon and gadid seqs to add to custom ref db"
output: html_document
date: "2024-12-27"
---

libraries
```{r}
library(tidyverse)
rename <- dplyr::rename
```

load the sample IDs 
```{r}
ids <- read.csv("/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/gadid_salmon_mifish_refdb.csv")
```


load asv table
```{r}
asv_table <- read_rds("/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqTab.RDS") %>%
  select(!Row.names)

#transpose 
asv_table <- data.frame(t(asv_table))

#set column names to be ASV# 
colnames(asv_table) <- asv_table["ASV",]

#remove row that has ASV#
asv_table <- asv_table[!rownames(asv_table) %in% c('ASV'), ]

#make sure reads are numbers
# Convert all character columns to numeric
for (col in names(asv_table)) {
  asv_table[[col]] <- as.numeric(asv_table[[col]])
}

#make make sample ID a column 
asv_table$Sample_ID <- rownames(asv_table)

asv_table <- asv_table %>%
  mutate(across(everything(), ~ gsub("-", "_", .))) %>%
  select(Sample_ID, everything()) 
```

join
```{r}
asv_table <- asv_table %>%
  left_join(ids, by = "Sample_ID") %>%
  select(Sample_ID, taxon, everything()) %>%
  mutate(across(-(1:2), as.numeric))
```

# Walleye pollock

pull out species that need seqs to be added to the ref db 
```{r}
wp <- asv_table %>%
  filter(taxon == "Gadus chalcogrammus") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique to pollock?
```{r}
wp_asvs <- wp %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(wp_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_01:ASV_33, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_wp = `Gadus chalcogrammus`/total)
```

```{r}
wp_filtered <- temp2 %>%
  filter(prop_wp > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta to be just the pollock
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/gadus_chalcogrammus.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
wp_seqs <- fasta_seqs[names(fasta_seqs) %in% wp_filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(wp_seqs, filepath = output_file)
```


# Pacific cod

pull out species that need seqs to be added to the ref db 
```{r}
pc <- asv_table %>%
  filter(taxon == "Gadus macrocephalus") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
pc_asvs <- pc %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(pc_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_04:ASV_32, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Gadus macrocephalus`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta 
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/gadus_macrocephalus.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```

# Polar cod

pull out species that need seqs to be added to the ref db 
```{r}
polc <- asv_table %>%
  filter(taxon == "Arctogadus glacialis") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
polc_asvs <- polc %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(polc_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_05:ASV_35, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Arctogadus glacialis`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/arctogadus_glacialis.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```

# Arctic cod

pull out species that need seqs to be added to the ref db 
```{r}
ac <- asv_table %>%
  filter(taxon == "Boreogadus saida") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
ac_asvs <- ac %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(ac_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_01:ASV_28, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Boreogadus saida`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta 
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/boreogadus_saida.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```

# Saffron cod

pull out species that need seqs to be added to the ref db 
```{r}
sc <- asv_table %>%
  filter(taxon == "Eleginus gracilis") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
sc_asvs <- sc %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(sc_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_12:ASV_24, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Eleginus gracilis`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta 
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/eleginus_gracilis.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```


# Tomcod

pull out species that need seqs to be added to the ref db 
```{r}
tc <- asv_table %>%
  filter(taxon == "Microgadus proximus") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
tc_asvs <- tc %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(tc_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_07:ASV_16, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Microgadus proximus`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta 
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/microgadus_proximus.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```



# Pink salmon

pull out species that need seqs to be added to the ref db 
```{r}
ogor <- asv_table %>%
  filter(taxon == "Oncorhynchus gorbuscha") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
ogor_asvs <- ogor %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(ogor_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_01:ASV_26, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Oncorhynchus gorbuscha`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta 
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/oncorhynchus_gorbuscha.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```


# Chum salmon

pull out species that need seqs to be added to the ref db 
```{r}
oketa <- asv_table %>%
  filter(taxon == "Oncorhynchus keta") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
oketa_asvs <- oketa %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(oketa_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_03:ASV_20, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Oncorhynchus keta`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta 
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/oncorhynchus_keta.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```


# Sockeye salmon

pull out species that need seqs to be added to the ref db 
```{r}
oner <- asv_table %>%
  filter(taxon == "Oncorhynchus nerka") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique?
```{r}
oner_asvs <- oner %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(oner_asvs)) %>%
  group_by(taxon) %>%
  summarise(across(ASV_02:ASV_27, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop = `Oncorhynchus nerka`/total)
```

```{r}
filtered <- temp2 %>%
  filter(prop > 0.1) %>%    ## at least 10% of reads come from the target species
  filter(total > 100) %>%          ## at least 100 reads 
  rownames()
```

subset fasta
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/data/building_custom_gadid_salmon_db/filtered.seqs.fasta"
output_file <- "/home/kimberly.ledger/GOA2021_mifish_mb/outputs/fastas_for_custom_db/oncorhynchus_nerka.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
seqs <- fasta_seqs[names(fasta_seqs) %in% filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(seqs, filepath = output_file)
```


## closer looks at ASV4 
pull out species that need seqs to be added to the ref db 
```{r}
temp <- asv_table %>%
  filter(taxon == "Gadus chalcogrammus" | taxon == "Gadus macrocephalus") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))

gadus_asvs <- temp %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(1, 2, gadus_asvs) %>%
  mutate(across(c(4:20), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_asv4 = ASV_04/total) %>%
  filter(prop_asv4 > 0.01)

asv_table %>%
  filter(taxon == "Gadus chalcogrammus")

asv_table %>%
  filter(taxon == "Gadus macrocephalus")
```

2/16 wp have asv 4 and only 1 has it as the dominate ASV (15285) - this individual was collected in Little Port Walter along with other pollock - some of which were included in this test and all came back as pollock. 

8/16 pc have asv 4 and they are all the dominate ASV



add ASV 19 and ASV20 for ref db as O. keta 
add ASV 18 and ASV 22 to ref db for O. gorbuscha




